---
title: "figure 2 and S2"
author: "MW"
date: "2024-01-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, echo=FALSE}
library(phyloseq)
library(eulerr)
library(VennDiagram)
library(ggnewscale)
library(venn)
library(grid)
library(gridExtra)
library(speedyseq)
library(DECIPHER)
library(ComplexHeatmap)
source("helper_functions.R")
source("decipher_idclusters.R")
```

Here is the color scheme used. 
```{r aesthetics}
origin_colors = c("mouth" = "#bc4b51", "crop" = "#f4a259", "nectar" = "#648FFF")
```

We begin by clustering ASVs to OTU at 97% identity to make sure we see the same trends even when we allow for more variability. Here we used the methods described in <https://github.com/benjjneb/dada2/issues/947>
```{r OTU_clusters, echo=FALSE}
# Set the number of cpus for clustering
set.seed(1279466746)
nproc <- 6
dna <- refseq(cmn_ps)
aln <- AlignSeqs(dna, processors = nproc)
d <- DistanceMatrix(aln, processors = nproc)
clusters <- IdClusters(d, method = "complete", cutoff = 0.03, processors = nproc)
cmn_OTU <- merge_taxa_vec(cmn_ps, group = clusters$cluster)
cmnOTU <- taxa_wrangle(cmn_OTU)
```

Comparing the number of OTUs in the winter season for each sample type. 
```{r winterVennD}
cmnOTU@sam_data$origin <- factor(cmnOTU@sam_data$origin, levels = c("mouth", "crop", "nectar"))
winTVL <- MicrobiotaProcess::get_vennlist(obj = subset_samples(cmnOTU, season == "winter"), "origin")
winVD = venn.diagram(winTVL, height = 5, width = 5, filename = NULL, fill = origin_colors[c(1,2,3)], cat.col = c(NA, NA, NA), 
                     alpha = 0.75, fontfamily = "serif", fontface = "plain", cex = 1.2, cat.cex = 1.3, 
                     cat.default.pos = "text", col = "black", cat.dist = 0.1, margin = 0.1, lwd = 2, 
                     lty = "solid", euler.d = F, scaled = F, inverted = T)
winVD_leg = Legend(at = c("mouth", "crop", "nectar"), title = "Sample type", legend_gp = gpar(fill = origin_colors[c(1,2,3)]))
grid.draw(winVD)
```

Plot Figure S2. Check to see that this pattern holds true for each *Apis* species. 
```{r beeVennD}
ceranaTVL <- MicrobiotaProcess::get_vennlist(obj = subset_samples(cmnOTU, bee_species == "cerana" & season == "winter"), "origin")
ceranaVD = venn.diagram(ceranaTVL, height = 5, width = 5, filename = NULL, fill = origin_colors[c(1,2,3)], cat.col = c(NA, NA, NA), 
                     alpha = 0.75, fontfamily = "serif", fontface = "plain", cex = 1.2, cat.cex = 1.3, cat.default.pos = "text", 
                     col = "black", cat.dist = 0.1, margin = 0.1, lwd = 2, lty = "solid", euler.d = F, scaled = F, 
                     inverted = T)

melliferaTVL <- MicrobiotaProcess::get_vennlist(obj = subset_samples(cmnOTU, bee_species == "mellifera" & season == "winter"), "origin")
melliferaVD = venn.diagram(melliferaTVL, height = 5, width = 5, filename = NULL, fill = origin_colors[c(1,2,3)], cat.col = c(NA, NA, NA), 
                     alpha = 0.75, fontfamily = "serif", fontface = "plain", cex = 1.2, cat.cex = 1.3, cat.default.pos = "text", 
                     col = "black", cat.dist = 0.1, margin = 0.1, lwd = 2, lty = "solid", euler.d = F, scaled = F, 
                     inverted = T)
grid.arrange(ceranaVD, melliferaVD, nrow = 1)
```

Jpeg file was created with the following code: 
```{r fig2_jpg}
top.vp = viewport(layout = grid.layout(3, 3, widths = unit(c(26, 160, 26), "mm"), heights = unit(c(26, 160, 26), "mm")))
plot.vp = viewport(layout.pos.col = 2, layout.pos.row = 2, name = "plot")
splot = vpTree(top.vp, vpList(plot.vp))

jpeg(filename = "Fig2.jpg", width = 238, height = 238, units = "mm", pointsize = 16, res = 400)
pushViewport(splot)
seekViewport("plot")
grid.draw(winVD)
draw(winVD_leg, x = unit(0.9, "npc"), y = unit(0.28, "npc"), just = c("right", "bottom"))
popViewport()
dev.off()
```




