---
title: "Figure 3 and S4"
author: "MW"
date: "2024-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages_and_setup, echo=FALSE}
options(dplyr.summarise.inform = FALSE)
library(phyloseq)
library(tidyverse)
library(vegan)
library(lme4)
library(microbiome)
source("helper_functions.R")
```

This script produces the alpha diversity Figures 3 and S4 in the manuscript. We made a data frame that included sample metadata as well as various alpha diversity estimates. Since we sampled the Nishi-honjo site twice, we labeled the summer sampling as Nishi-honjo_S site and the winter sampling as Nishi-honjo_W site. 
```{r alpha_div}
alpha_measures = c("Shannon", "Observed")
cmn_alpha <- data.frame(cmnOTU@sam_data, estimate_richness(cmnOTU, measures = alpha_measures)) %>% 
  mutate(site2 = ifelse(site == "Nishi-honjo", ifelse(season == "winter", "Nishi-Honjo_W", "Nishi-Honjo_S"), site), 
         site_alpha = factor(site2, levels = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S","Nishi-Honjo_W", "Higashi-honjo", "Hirono", "Kamihaya", "Kiyokawa 2", "Nagano", "Oshine 2", "Shimanose", "Yamauchi")), 
         origin = factor(origin, levels = c("crop", "mouth", "nectar")))
```

Alpha diversity and evenness statistics. 
```{r stats}
cmn_alpha = merge(cmn_alpha, microbiome::evenness(cmnOTU, index = "pielou"), by = 0) %>% 
  mutate(pielou = ifelse(is.nan(pielou), 0, pielou)) %>% 
  column_to_rownames(var = "Row.names")


# Filter for winter season and compare crop to mouth, mouth to nectar, and nectar to crop. Filter for summer season and compare crop to mouth
alpha_stats <- lapply(c("summer", "winter"), function(s) {
  df <- cmn_alpha %>% filter(season == s)
  combos <- combn(unique(df$origin), 2)
  stats <- lapply(seq_len(ncol(combos)),function(c) {
    Shannon_stat <- t.test(df[df$origin == combos[,c][2], "Shannon"],
                           df[df$origin == combos[,c][1], "Shannon"], alternative = "two.sided")
    even_stat <- t.test(df[df$origin == combos[,c][2], "pielou"],
                        df[df$origin == combos[,c][1], "pielou"], alternative = "two.sided")
    return(list("shannon" = Shannon_stat, "evenness" = even_stat))
  })
  names(stats) <- paste(combos[2,], combos[1,], sep = "_")
  df <- df %>% group_by(origin) %>% summarize(shannon_mean = mean(Shannon), evenness_mean = mean(pielou))
  return(list("stats" = stats, "means" = df))
})
names(alpha_stats) <- c("summer", "winter")
alpha_stats
```


Here are the color and shape schemes used. 
```{r aesthetics}
origin_colors = c("crop" = "#f4a259", "mouth" = "#bc4b51", "nectar" = "#648FFF")
origin_shapes = c("crop" = 25, "mouth" = 23, "nectar" = 22)
bee_shapes = c("cerana" = 16, "mellifera" = 17)
```

For the bootstrapping function we are interested in the random effect of site. 
```{r bootfun}
bootfun <- function(x) c(as.numeric(ranef(x)[["site_alpha"]][,1]), fixef(x)[["(Intercept)"]])
```

We also want to compare the samples within season. 
```{r sample_season_combos}
sample_type = c("crop", "mouth", "nectar")
season = c("winter", "summer")
ss_comb = crossing(sample_type, season) %>% 
  rowwise() %>% 
  mutate(comm = list(c(sample_type, season)))
ss_comb <- append(ss_comb$comm[c(1:4,6)], sample_type, after = 0)
```

There was no nectar to collect from the Japanese apricot flowers in the summer, so to make the plot create empty spaces in the summer portion of the nectar column we an empty summary data frame for nectar. 
```{r summer_nectar}
pseudo_nec = data.frame(site = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S"), season = rep("summer", 4), 
           mean = rep(NA, 4), med = rep(NA, 4), lo = rep(NA, 4), hi = rep(NA, 4))
pseudo_summer_nectar = data.frame(origin = rep("nectar", 4), sample_name = rep(NA, 4), date = rep(NA, 4), 
                                  season = rep("summer", 4), year = rep(NA, 4), bee_genera = rep(NA, 4), 
                                  bee_species = rep(NA, 4), pollen_basket = rep(NA, 4), 
                                  location = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S"), 
                                  Lk_like_CFU_crop = rep(NA, 4), Pa_like_CFU_crop = rep(NA, 4), elevation = rep(NA, 4), 
                                  longitude = rep(NA, 4), latitude = rep(NA, 4), 
                                  site = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S"), 
                                  Observed = rep(NA, 4), Shannon = rep(NA, 4), 
                                  site2 = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S"), 
                                  site_alpha = c("Ichiigawa", "Kiyokawa 1", "Oshine 1", "Nishi-Honjo_S"), 
                                  pielou = rep(NA, 4))
```

Next we bootstrap the alpha diversity model with site as a random effect, and create plots of both the alpha diversity and random effect coefficient for each alpha diversity measure. 
```{r alpha_bootstrap}
set.seed(1279466746)
alpha_sim = lapply(ss_comb, function(tissue) {
    alpha_df <- alpha_wrangle(cmn_alpha, tissue)
    results = lapply(alpha_measures, function(m) {
        alpha_res <- boot_res(alpha_measure = m, bootfun = bootfun, df = alpha_df)
        ref_df <- random_effect_df(alpha_res$results)
        rich_df <- alpha_results_df(alpha_res$results)
        
        if (unique(alpha_df$origin) == "nectar") {
          ref_df <- rbind(ref_df, pseudo_nec)
          rich_df <- rbind(rich_df, pseudo_nec)
          alpha_df <- rbind(alpha_df, pseudo_summer_nectar) 
        }
        
        ref_plot <- random_effect_plot(ref_df, tissue)
        alpha_plot <- alpha_results_plot(rich_df, alpha_df, tissue, m)
        
        return(list("lm" = alpha_res$alpha_mod, "sim" = alpha_res$results, "ranef_plot" = ref_plot, "alpha_plot" = alpha_plot))
  })
  names(results) = alpha_measures
  return(results)
})
names(alpha_sim) = c("crop", "mouth", "nectar", "summer_crop", "winter_crop", "summer_mouth", 
                     "winter_mouth", "winter_nectar")
```

The code below creates Figure 3. 
```{r fig3}
# Save 8 x 4 inches
grid.arrange(alpha_sim$crop$Shannon$alpha_plot + ylim(-0.2, 4) + theme_classic() + 
                 theme(axis.title = element_blank(), axis.text.x = element_blank(), 
                       axis.ticks.x = element_blank(), legend.position = "none"), 
             alpha_sim$mouth$Shannon$alpha_plot + ylim(-0.2, 4) + theme_classic() + 
                 theme(axis.title = element_blank(), axis.ticks.x = element_blank(), 
                       axis.text.x = element_blank(), legend.position = "none"), 
             alpha_sim$nectar$Shannon$alpha_plot + ylim(-0.2, 4) + theme_classic() + 
                 theme(axis.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), 
                       legend.position = "none"), 
             textGrob("Shannon", rot = 90, gp = gpar(col = "black", fontsize = 16)),
             layout_matrix = rbind(c(NA, 1), c(4, 2), c(NA, 3)), 
             heights = c(0.3, 0.3, 0.4), 
             widths = c(0.04, 0.96))
```

The code below creates Figure S4. 
```{r figS4}
grid.arrange(alpha_sim$crop$Shannon$ranef_plot + xlim(-1.5, 2.5) + theme_classic() + 
               theme(axis.title.x = element_blank()), 
             alpha_sim$mouth$Shannon$ranef_plot + xlim(-1.5, 2.5) + theme_classic() + 
               theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
                     axis.title.y = element_blank()), 
             alpha_sim$nectar$Shannon$ranef_plot + xlim(-1.5, 2.5) + theme_classic() + 
               theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
                     axis.title = element_blank()), 
             layout_matrix = rbind(c(1, 2, 3), c(1, 2, 3), c(NA, 2, NA)), 
             heights = c(0.255, 0.65, 0.05), 
             widths = c(0.44, 0.28, 0.28))
```


